


## sketch

  - stan code for a random-intercept with no latent groups model
  - update said model to have latent groups

Simulate some repeated measures data. There are no latent group effects yet.
Happy with the idea of storing stuff in dataframe and joining at the
end, not happy with the verbosity.

```{r}
library(tidyverse)
rep_along <- function (along, x) rep_len(x, length(along))

simulate_data <- function(
  n_individuals = 20,
  n_obs = 100,
  n_groups = 3,
  sigma_y = .2,
  group_mean = NULL,
  group_sigma = NULL
) {
  if (is.null(group_mean)) {
    group_mean <- rep_len(0, n_groups)
    group_sigma <- rep_len(1, n_groups)
  }
  if (is.null(group_sigma)) {
    group_sigma <- rep_len(1, n_groups)
  }
  
  d_groups <- data.frame(
    group = seq_len(n_groups),
    group_mean = group_mean,
    group_sigma = group_sigma
  )
  
  d_observations <- data.frame(
    y = rep_len(NA, n_obs),
    individual = NA_integer_
  )
  
  d_individuals <- data.frame(
    individual = seq_len(n_individuals),
    group = NA_integer_,
    individual_mean = NA_real_
  )

  d_individuals$group <- sample(
    seq_len(n_groups), 
    n_individuals, 
    replace = TRUE
  )

  d_observations$individual <- sample(
    d_individuals$individual, 
    n_obs, 
    replace = TRUE
  )

  d_individuals$individual_mean <- rnorm(
    n_individuals,
    d_groups$group_mean[d_individuals$group],
    d_groups$group_sigma[d_individuals$group]
  )

  d_observations$y <- rnorm(
    n_obs,
    d_individuals$individual_mean[d_observations$individual], 
    sigma_y
  )

  d_observations |> 
    left_join(d_individuals, by = join_by(individual)) |> 
    left_join(d_groups, by = join_by(group)) |> 
    mutate(sigma_y = sigma_y)
}

d <- simulate_data()

```

Run a simple Stan model and check that it recovers the parameters.

```{r simple-ri-model, fig.width = 6, fig.height = 3, out.width="80%"}
m <- cmdstanr::cmdstan_model("1.stan")
data <- list(
  n_obs = nrow(d),
  n_ind = length(unique(d$individual)),
  individual = d$individual,
  n_groups = length(unique(d$group)),
  y = d$y
)
e <- m$sample(data, refresh = 0)
e_sum <- e$summary()

d_post_individuals <- e_sum |> 
  filter(variable |> startsWith("coef_b[")) |> 
  mutate(
    individual = readr::parse_number(variable),
    type = "alpha + ranef"
  ) |> 
  left_join(
    d |> distinct(individual, individual_mean), 
    by = join_by(individual)
  )

ggplot(d_post_individuals) + 
  aes(x = individual) + 
  geom_linerange(
    aes(ymin = q5, ymax = q95, color = "posterior 90%\n(inferred latent mean)"), 
  ) +
  geom_point(
    aes(y = individual_mean, color = "latent mean"),
    size = 3,
    alpha = .8
  ) + 
  geom_point(
    aes(y = y, color = "observed mean"),
    data = d,
    alpha = .8,
    stat = "summary",
    size = 3,
  ) +
  guides(
    color = guide_legend(
      override.aes = list(shape = c(19, 19, NA), linewidth = c(NA, NA, .5))
    )
  ) + 
  facet_wrap("type") +
  scale_color_manual(
    "quantities",
    values = palette.colors(3)[c(2, 3, 1)] |> unname()
  )
```

```{r simple-ri-model-2, fig.height = 3, fig.width = 4}
e_sum |> 
  filter(variable %in% c("alpha", "sigma_b", "sigma_y")) |> 
  left_join(
    d |> 
      distinct(group_mean, group_sigma, sigma_y) |> 
      rename(sigma_b = group_sigma, alpha = group_mean) |> 
      tidyr::pivot_longer(
        everything(), 
        names_to = "variable", 
        values_to = "value"
      )
  ) |> 
  ggplot() + 
    aes(x = variable) + 
    geom_linerange(aes(ymin = q5, ymax = q95)) +
    geom_point(
      aes(y = value),
      color = "blue",
      size = 3,
      alpha = .3
    )
```


## Assignment to groups

Let's assume the individuals belong to latent groups. 
```{r latent-groups, out.width = "80%", fig.width = 6, fig.height = 3}
d <- simulate_data(
  n_groups = 3, 
  group_mean = c(-3, 1, 3.5)
)

ggplot(d) + 
  aes(x = individual, y = y) + 
  geom_point(aes(color = factor(group))) +
  geom_point(aes(y = individual_mean))

```

The random intercept model does okay.

```{r simple-ri-model-3, fig.width = 6, fig.height = 3, out.width="80%"}
m <- cmdstanr::cmdstan_model("1.stan")

data <- list(
  n_obs = nrow(d),
  n_ind = length(unique(d$individual)),
  individual = d$individual,
  n_groups = length(unique(d$group)),
  y = d$y
)
e <- m$sample(data, refresh = 0)
e_sum <- e$summary()

d_post_individuals <- e_sum |> 
  filter(variable |> startsWith("coef_b[")) |> 
  mutate(
    individual = readr::parse_number(variable),
    type = "alpha + ranef"
  ) |> 
  left_join(
    d |> distinct(individual, individual_mean), 
    by = join_by(individual)
  )

ggplot(d_post_individuals) + 
  aes(x = individual) + 
  geom_linerange(
    aes(ymin = q5, ymax = q95, color = "posterior 90%\n(inferred latent mean)"), 
  ) +
  geom_point(
    aes(y = individual_mean, color = "latent mean"),
    size = 3,
    alpha = .8
  ) + 
  geom_point(
    aes(y = y, color = "observed mean"),
    data = d,
    alpha = .8,
    stat = "summary",
    size = 3,
  ) +
  guides(
    color = guide_legend(
      override.aes = list(shape = c(19, 19, NA), linewidth = c(NA, NA, .5))
    )
  ) + 
  facet_wrap("type") +
  scale_color_manual(
    "quantities",
    values = palette.colors(3)[c(2, 3, 1)] |> unname()
  )
```

```{r, eval = FALSE}
m <- cmdstanr::cmdstan_model("2.stan")
data <- list(
  n_obs = nrow(d),
  n_ind = length(unique(d$individual)),
  individual = d$individual,
  n_groups = length(unique(d$group)),
  y = d$y
)
e <- m$sample(data, refresh = 0)
e_sum <- e$summary()
e_sum |> print(n = Inf)
```
