---
title: "Latent class mixed effects models"
bibliography: refs.bib
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE, 
  dpi = 300,
  out.width = "40%"
)
library(tidyverse)

```

## Background

@JSSv078i02 describe the statistical machinery for their latent class
mixed models. We note the following equations from their paper.

Below is the likelihood contribution for one individual in the basic
linear mixed model that we know and love:

$$
\displaylines{
L_i = \phi_i(Y_i; \theta_1) \\
\phi: \textrm{MVN density} \\
i: \textrm{individual index}
}
$$

When there are $G$ latent classes, the likelihood becomes a weighted sum
of class-specific likelihoods:

$$
\displaylines{
L_i(\theta_G) = \sum_{g=1}^G \pi_{ig}\phi_{ig}(Y_i|c_i = g; \theta_G) \\
\pi_{ig}: \textrm{probability of group membership for an individual} \\
g: \textrm{group index}
}
$$


So, an individual makes $G$ contributions to the likelihood and each one
is weighted by their group membership probability. And when there is one
group, this equation reduces to the first likelihood equation.

Finally, group probabilities are defined as a multinomial logistic
model:

$$
\displaylines{
\pi_{ig} = 
  P(c_i = g | X_{ci}) = 
  \frac{
    e^{\xi_{0g} +X^{\top}_{ci}\xi_{1g}}
  }{
    \sum^{G}_{l=1}e^{\xi_{0l} +X^{\top}_{ci}\xi_{1l}}
  } \\
c_i : \textrm{the latent class for an individual} \\
X_{ci} : \textrm{time-indpendent covariates} \\
\xi_{0g} +X^{\top}_{ci}\xi_{1g} : \textrm{linear model for group membership}
}
$$

## sketch

  - stan code for a random-intercept with no latent groups model
  - update said model to have latent groups

Simulate some repeated measures data. There is no latent group effects.
Happy with the idea of storing stuff in dataframe and joining at the
end, not happy with the verbosity.

```{r}
library(tidyverse)
rep_along <- function (along, x) rep_len(x, length(along))

n_individuals <- 20
n_obs <- 100
n_groups <- 3
sigma_y <- .2

d_groups <- data.frame(
  group = seq_len(n_groups),
  group_mean = 0,
  group_sigma = 1
)

d_observations <- data.frame(
  y = rep_len(NA, n_obs),
  individual = NA_integer_
)

d_individuals <- data.frame(
  individual = seq_len(n_individuals),
  group = NA_integer_,
  individual_mean = NA_real_
)

d_individuals$group <- sample(
  seq_len(n_groups), 
  n_individuals, 
  replace = TRUE
)

d_observations$individual <- sample(
  d_individuals$individual, 
  n_obs, 
  replace = TRUE
)

d_individuals$individual_mean <- rnorm(
  n_individuals,
  d_groups$group_mean[d_individuals$group],
  d_groups$group_sigma[d_individuals$group]
)

d_observations$y <- rnorm(
  n_obs,
  d_individuals$individual_mean[d_observations$individual], 
  sigma_y
)

list(d_groups, d_individuals, d_observations) |> lapply(head)

d <- d_observations |> 
  left_join(d_individuals, by = join_by(individual)) |> 
  left_join(d_groups, by = join_by(group))
```

Run a simple stan model and check that it recovers the parameters.

```{r simple-ri-model, fig.width = 3, fig.height = 3}
m <- cmdstanr::cmdstan_model("1.stan")

data <- list(
  n_obs = n_obs,
  n_ind = n_individuals,
  individual = d$individual,
  n_groups = n_groups,
  y = d$y
)
e <- m$sample(data, refresh = 0)
e_sum <- e$summary()

d_post_individuals <- e_sum |> 
  filter(variable |> startsWith("b")) |> 
  mutate(individual = readr::parse_number(variable)) |> 
  left_join(d_individuals, by = join_by(individual))

ggplot(d_post_individuals) + 
  aes(x = individual) + 
  geom_linerange(aes(ymin = q5, ymax = q95)) +
  geom_point(
    aes(y = individual_mean),
    color = "blue",
    size = 3,
    alpha = .3
  )

e_sum |> 
  filter(variable %in% c("alpha", "sigma_b", "sigma_y")) |> 
  left_join(
    tibble(
      variable = c("alpha", "sigma_b", "sigma_y"),
      value = c(0, 1, sigma_y)
    )
  ) |> 
  ggplot() + 
    aes(x = variable) + 
    geom_linerange(aes(ymin = q5, ymax = q95)) +
    geom_point(
      aes(y = value),
      color = "blue",
      size = 3,
      alpha = .3
    )

```




## A random intercept model

Each group has their own mean.

```{r, fig.width = 3, fig.height = 3, eval = FALSE}

ggplot(d) + 
  aes(x = i, y = y) + 
  geom_point(aes(color = factor(ci))) +
  stat_smooth(
    aes(color = factor(ci)),
    data = function(x) x |> group_by(ci, i) |> summarise(y = mean(y)),
    method = "lm",
    formula = y ~ 1
  )

d |> 
  group_by(ci) |> 
  summarise(mean(y), sd(y))

# means[]
# rnorm()

means[assignments]
sigmas[assignments]

```

