---
title: "Latent class mixed effects models"
bibliography: refs.bib
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

@JSSv078i02 describe the statistical machinery for their latent class mixed
models. We note the following equations from their paper.

Below is the likelihood contribution for one individual in the basic linear
mixed model that we know and love:

$$
\displaylines{
L_i = \phi_i(Y_i; \theta_1) \\
\phi: \textrm{MVN density} \\
i: \textrm{individual index}
}
$$

When there are $G$ latent classes, the likelihood becomes a weighted sum of
class-specific likelihoods:

$$
\displaylines{
L_i(\theta_G) = \sum_{g=1}^G \pi_{ig}\phi_{ig}(Y_i|c_i = g; \theta_G) \\
\pi_{ig}: \textrm{probability of group membership for an individual} \\
g: \textrm{group index}
}
$$


So, an individual makes $G$ contributions to the likelihood and each one is
weighted by their group membership probability. And when there is one group,
this equation reduces to the first likelihood equation.

Finally, group probabilities are defined as a multinomial logistic model:

$$
\displaylines{
\pi_{ig} = P(c_i = g | X_{ci}) = \frac{e^{\xi_{0g} +X^{\top}_{ci}\xi_{1g}}}{\sum^{G}_{l=1}e^{\xi_{0l} +X^{\top}_{ci}\xi_{1l}}}
} \\
c_i : \textrm{the latent class for an individual} \\
X_{ci} : \textrm{time-indpendent covariates} \\
\xi_{0g} +X^{\top}_{ci}\xi_{1g} : \textrm{linear model for group membership}

$$

## sketch

- stan code for a random-intercept with no latent groups model
- update said model to have latent groups

```{r, eval=FALSE}
m <- cmdstanr::cmdstan_model("1.stan")

d <- list(
  n_obs = n_obs,
  n_ind = length(individuals),
  jj = observation_to_individual,
  n_groups = groups,
  y = observations
)
e <- m$sample(d)
e_sum <- e$summary()

e_sum |> 
  filter(variable |> startsWith("b")) |> 
  mutate(variable = readr::parse_number(variable)) |> 
  ggplot() + 
  aes(x = variable) + 
  geom_pointrange(aes(y = mean, ymin = q5, ymax = q95)) +
  geom_point(
    aes(y = y),
    tibble(y = observations - mean(observations), variable = observation_to_individual),
    color = "blue"
  )
e_sum |> tail()

```




## A random intercept model

Each group has their own mean.

```{r, fig.width = 3, fig.height = 3}
individuals <- 1:40
n_obs <- 200
groups <- 3
overall_sigma <- 1

individual_to_group <- sample(seq_len(groups), length(individuals), replace = TRUE)
observation_to_individual <- sample(individuals, n_obs, replace = TRUE)
observation_to_group <- individual_to_group[observation_to_individual]

group_population_means <- seq_len(groups) * 2
group_population_sigmas <- rep(.5, groups)

individual_means <- rnorm(
  length(individuals),
  group_population_means[individual_to_group],
  group_population_sigmas[individual_to_group]
)

observations <- rnorm(
  n_obs,
  individual_means[observation_to_individual], 
  overall_sigma
)

library(tidyverse)
d <- tibble(
  y = observations,
  i = observation_to_individual,
  ci = observation_to_group
)

ggplot(d) + 
  aes(x = i, y = y) + 
  geom_point(aes(color = factor(ci))) +
  stat_smooth(
    aes(color = factor(ci)),
    data = function(x) x |> group_by(ci, i) |> summarise(y = mean(y)),
    method = "lm",
    formula = y ~ 1
  )

d |> 
  group_by(ci) |> 
  summarise(mean(y), sd(y))

# means[]
# rnorm()

means[assignments]
sigmas[assignments]

```

